Staged Abstract Interpreters

### 2019-09-04 paper/10_intro.tex
Proofreading intro

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST). The traversing incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation. 
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.  
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster. The Futamura projections has been successfully applied to 
other similar performance problems, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275} and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-point iteratively over a cache. The solution is not simply obvious.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model.
In short, our approach borrows ideas and techniques from multi-stage
programming and embedded DSLs to abstract interpreters:  1) To remove the
overhead from interpretation and effect layers, we specialize the abstract
interpreter via staging and then generate efficient low-level code that does
the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times and different semantics, which allows the staged
abstract interpreter can be derived from its unstaged counterpart.
As a result, the derived staged abstract interpreters has no intrusive changes
to the underlying abstract semantics, therefore preserves the soundness.
In this sense, our approach allows for no regret for both performance and
engineering effort. We elaborate these two main ideas in detail, as follows.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

@@

Program specialization and abstract interpretation are two orthogonal
concepts.  To implement the confluence of them, we first construct a
generic interpreter that is agnostic to both binding-times and value
domains used in the semantics.  Later, the generic interpreter can be
instantiated from these two dimensions (Figure~\ref{confluence}):
\begin{itemize}
\item With a flat binding-time and concrete domains, it is an ordinary
  definitional interpreter based on big-step operational semantics;
\item With two-level binding-times and concrete domains, it is a
  compiler that translates a program into another language;
\item With a flat binding-time and abstract domains, it is a
  definitional abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}
  that statically computes runtime properties;
\item With two-level binding-times and abstract domains, it is an optimizing
  program analyzer, but works in the fashion of compilation.
\end{itemize}

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, except the stage annotations, 
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup on flow analyses.

>>>>

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST). The traversal incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation.
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. The Futamura projections have been successfully applied to
other similar performance problems, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275}, and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively over a cache. The solution is not simply obvious.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model as possible.
In short, our approach borrows ideas and techniques from multi-stage
programming and embedded DSLs to abstract interpreters:  1) To remove the
overhead from interpretation and effect layers, we specialize the abstract
interpreter via staging and then generate efficient low-level code that does
the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times and different semantics, which allows the staged
abstract interpreter to be derived from its unstaged counterpart.
As a result, the derived staged abstract interpreters has no intrusive changes
to the underlying abstract semantics, thereby preserving the soundness.
In this sense, our approach allows for no regret for both performance and
engineering effort. We elaborate these two main ideas in detail, as follows.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

@@

Program specialization and abstract interpretation are two orthogonal
concepts.  To implement the confluence of them, we first construct a
generic interpreter that is agnostic to both binding times and value
domains used in the semantics.  Later, the generic interpreter can be
instantiated from these two dimensions (Figure~\ref{confluence}):
\begin{itemize}
\item With a flat binding-time and concrete domains, it is an ordinary
  definitional interpreter based on big-step operational semantics;
\item With two-level binding-times and concrete domains, it is a
  compiler that translates a program into another language;
\item With a flat binding-time and abstract domains, it is a
  definitional abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}
  that statically computes runtime properties;
\item With two-level binding-times and abstract domains, it is an optimizing
  program analyzer, but works in the fashion of compilation.
\end{itemize}

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, except the addition of stage annotations,
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup on flow analyses.
